---
title: "Climate Change Module"
author: 'Add name'
format:
  html:
    embed-resources: true
---

```{r message=FALSE}
library(tidyverse)
library(ggthemes)
```

## Warm-up: Examining CO2 trends in R

-   Example from <http://climate.nasa.gov/vital-signs/carbon-dioxide/>
-   Raw data from <https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt>

In 1958, Dr. Charles David Keeling (1928-2005), a scientist at Scripps Institute of Oceanography, began collecting data on atmospheric CO2 concentration at the Mauna Loa Observatory located in Hawaii. This dataset allowed us to understand the degree to which climate change is human-caused through our burning of fossil fuels and release of CO2 into the atmosphere. Due to his scientific achievements, Dr. Keeling was awarded the National Medal of Science by President George W. Bush in 2002. This is the highest award for lifetime scientific achievement that can be granted in the U.S. Today, you get to analyze this same dataset, except that you have more data than was available to Dr. Keeling and his colleagues because your dataset extends up to the current time.

To read the code, you will use a new function called `read_table`. It is similar to `read_csv` except it looks for spaces between column entries rather than commas (remember that csv stands for comma-separated values). Others include `read_tsv`, which uses tabs to separate entries. You can discover the separation type by putting <https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt> into your web browser and examining the file. The code also includes an argument for comment, which denotes the character used to define a line in the file as being a comment rather than data or a header. The file also reveals that the column headers are on lines with a comment character, so they won't be read. You will use the argument `col_names` to define the headers manually. Finally, `NA` values are defined by the numbers -1 and -99, so they need to be defined using the `na` argument (otherwise, they will be read as numeric values).

```{r message=FALSE}

co2 <-  read_table("https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt", 
                  comment="#",
                  col_names = c("year", "month", "decimal_date",
                                "monthly_average",
                                "deseasonalized", "days", "sd_days",
                                "unc_month_mean"),
                  na = c("-1", "-99.99"))
co2
```

```{r}
ggplot(co2, aes(x = decimal_date, y = monthly_average)) + 
  geom_line() + 
  geom_line(aes(y = deseasonalized), color = "blue") +
  labs(x = "Year", y = "CO2 concentration (ppm)")
```

**Question 1:**

Describe the overall trend in the CO2 data.

**Answer 1:**

The trend in the C02 concentration over the years is positive. The graph shows a slightly upward curved line, indicating the CO2 concentration has continue to grow throughout the years since 1958.

**Question 2:**

How does CO2 vary within a year? What month is it at max? Min? What might explain this sub-annual pattern? (you will need to write code and make a plot to answer this question)

**Answer 2:**

```{r}
ggplot(co2 |>  filter(year == 1962), 
       aes(x = month, y = monthly_average)) + 
  geom_line() + 
  scale_x_continuous(breaks = 1:12, labels = 1:12) + 
  labs(x = "Month", y = "CO2 concentration (ppm)", 
       title = "CO2 Concentration in 1958") +
  theme_fivethirtyeight()
```

In order to look at how to CO2 concentrations vary within a given year, I decided to make a graph of the CO2 levels in the year 1959 as it was the first year in the dataset with all 12 months. The year can be swapped out in the ggplot code in order to look at others year's levels. The peak CO2 levels in 1959 occurs in the month of May, and the minimum levels occur in October. When looking at different years and their respective CO2 levels per month, most years seem to also show an upward trend from January, until it peaks in May and then declines. As the summer months roll around, the levels decreases and are at their lowest in October. The levels then rise again in the following Winter months November and December. This sub-annual pattern could in part be explained by photosynthesis. During the Spring and Summer months, photosynthesis among plants is at its peak due to the newly flowered plants. Since CO2 is absorbed during photosynthesis, this could explain why the CO2 levels begin to decline in Spring and Summer, and rise in Winter and Fall (Not as much photosynthesis during winter results in more CO2 concentration levels).

## Global Temperature Data

Current climate change affects many aspects of the environment, with socio-economic consequences. For example, a warmer climate can allow new diseases to be introduced and persist (e.g., West Nile became established in the United States after an unusually warm winter, which allowed the mosquitoes that carry the virus to survive and spread). We are concerned not only with the actual temperature but also with the rate at which it changes. Rapid changes make it more likely that species cannot adapt and will go extinct.

Each of the most recent years has been the warmest on record. In this section, we will analyze global mean temperature data.

Data from: <https://climate.nasa.gov/vital-signs/global-temperature/>

**Question 3:**

Describe the data set to the best of your ability, given the documentation provided.

-   Where is the data from?
-   Describe the class of each column and what units it is measured in.
-   What is the difference between "global temperature" and "global temperature anomaly"?

**Answer 3:**

The data is gathered by NASA at the Goddard Institute for Space Studies (GISS). The year column in numeric and is simply the year the data was recorded in. No_Smoothing is a numeric and shows the actual mean global temperature in degrees celcius. Lowess_5 is a numeric and represents the mean global temperature in degrees celcius with a LOWESS (Locally Weighted Scatterplot Smoothing) degree of 5 for visual purposed and to better view the trend in the data. The difference between global temperature (recorded value per year) and the global temperature anomaly is that the anomaly shows the change in global surface temperature compared to the long-term average from 1951 to 1980. The dataset uses this period as a baseline average to view how the temperatures changed from 1880 to 2024 in reference to this average.

**Question 4:**

Construct the necessary R code to import and prepare for plotting the following data set: <https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt>

You'll need to determine the file's delimiter and any comments or skips. You will also need to be sure that you have column names. You will not directly use the code above to read the CO2 data, but that code provides helpful tips for reading the data.

**Answer 4:**

```{r message=FALSE}
climate <- read_table("https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt", 
                   skip = 5,
                   col_names = c("Year", "No_Smoothing", "Lowess_5"),
                   na = c("-1", "-99.99"))
climate
```

**Question 5:**

Plot the trend in global mean temperatures over time. Describe what you see in the plot and how you interpret the patterns you observe.

**Answer 5:**

```{r}
ggplot(data, aes(x = Year)) +
  geom_line(aes(y = No_Smoothing, color = "Annual Mean")) +
  geom_line(aes(y = Lowess_5, color = "Lowess Smoothing")) +
  labs(title = "Global Mean Temperature Trends",
       x = "Year", y = "Temperature Anomaly (°C)",
       color = "Legend") +
  theme_minimal()
```
When looking at the plot, it is clear that temperatures overall have been throughout the years since 1880. The first time the temperature anomaly reached 0 for the time in 1938, and it first became positive in 1940. In the early to mid 1960s, the temperature anomaly began a rise that continued until 2024. The 10 most recent years are the warmest years on record, which goes to show just how much the Earth has been heating up as of recently, and throughout history.

## Evaluating the evidence for a "Pause" in warming?

The [2013 IPCC Report](https://www.ipcc.ch/pdf/assessment-report/ar5/wg1/WG1AR5_SummaryVolume_FINAL.pdf) included a tentative observation of a "much smaller increasing trend" in global mean temperatures since 1998 than was observed previously. This led to much discussion in the media about the existence of a "Pause" or "Hiatus" in global warming rates, as well as much research looking into where the extra heat could have gone. (Examples discussing this question include articles in [The Guardian](http://www.theguardian.com/environment/2015/jun/04/global-warming-hasnt-paused-study-finds), [BBC News](http://www.bbc.com/news/science-environment-28870988), and [Wikipedia](https://en.wikipedia.org/wiki/Global_warming_hiatus)).

You will use rolling averages to help you explore the evidence for a pause. Since you have not been provided instructions for calculating rolling means, the learning objective of this question is to practice finding the solution.

**Question 6:**

Use a search engine (e.g., Google) or a chat LLM to find out how to calculate a rolling average in R. What search term or chat LLM prompt did you use? What website or answer from the chat LLM did you end up using? How much did you need to modify the code from the website or chat LLM to answer the question successfully?

**Answer 6:**

In order to calculate a rolling average in R, you must use the rollmean() function. I simply looked up on Google "how do you calculate a rolling average in R" and this was the first response under the "AI Overview" tab. It advised me that list function uses the "zoo" package and is formatted like so rollmean(your_data, k = window_size, fill = NA, align = "right"). I have to modify the code to include my data the desired window size, the fill, and the allignment. 

**Question 7:**

-   What is the meaning of "5-year average" vs. "annual average"?
-   Create a data frame from the annual temperature anomaly data (from Question 4) with three new columns: 5-year running averages, 10-year running averages, and 20-year running averages.

**Answer 7:**

The annual average is the average temperature anomaly for a single year. A 5-year average is the average temperature anomalies over a 5 year period, or window. 

```{r message=FALSE}
install.packages("zoo")
library(zoo)

climate <- read_table("https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt", 
                   skip = 5,
                   col_names = c("Year", "No_Smoothing", "Lowess_5"),
                   na = c("-1", "-99.99"))

climate$Rolling_5 <- rollmean(climate$No_Smoothing, k = 5, fill = NA, align = "right")
climate$Rolling_10 <- rollmean(climate$No_Smoothing, k = 10, fill = NA, align = "right")
climate$Rolling_20 <- rollmean(climate$No_Smoothing, k = 20, fill = NA, align = "right")

climate
```

**Question 8:**

Plot the different averages on the *same plot* and describe what differences you see and why.

**Answer 8:**

```{r}
ggplot(climate, aes(x = Year)) +
  geom_line(aes(y = Rolling_5, color = "5-Year Avg"), size = 1) +
  geom_line(aes(y = Rolling_10, color = "10-Year Avg"), size = 1) +
  geom_line(aes(y = Rolling_20, color = "20-Year Avg"), size = 1) +
  labs(title = "Global Average Temperature with Rolling Averages",
       x = "Year", y = "Temperature Anomaly (°C)",
       color = "Legend") +
  theme_minimal()
```

The first thing I notice about this graph above is that the 5-year average line seems to be much more jagged and there appears to be more fluctuation in the temperature anomalies than the 10 and 20-year averages. The 20-year average line appears much more rounded and smooth. This makes sense as the 20-year data is averaging 20 different yearly temperatures so there would be less variation in the points along the graph. This also explains why the 5-year average line appears much more jagged, as there is only a 5 year average which could more heavily factor in outliers, or trends in a short period.  

**Question 9:**

By examining the data here, what evidence do you find or not find for such a pause?

**Answer 9:**

When looking at this graph, in particular the 20-year rolling average line, it becomes clear that there is no pause, or hiatus in global warming. The temperature anomaly has been steadily increasing since the late 1960s and early 1970s. Seeing how smooth the 20-year average line is just shows how steadily the temperature has been increasing without pauses or breaks in the warming. Overall, there might be slight declines in average temperature per year due to weather anomalies or extreme years, but at the current rate we are progressing, there not seem to be any pauses, or breaks in global warming anytime soon. 


## Longer term trends in CO2 Records

When analyzing Earth’s climate, it is important to remember that Earth is 4.54 billion years old. Our analyses so far have only looked at recent history. How can we compare the recent data to prehistoric times? Are the current rates of change similar or different from those the earth has experienced in the past? To explore this, we can use data from ice cores drilled at the poles.

Hundreds of ice cores have been extracted from polar ice because they contain valuable data on atmospheric chemistry over pre-historic times. These valuable data exist in tiny air bubbles trapped in the ice. These air bubbles contain the same gases in the same ratios as the atmosphere at the time when the ice formed. The data you will analyze today are from ice cores extracted from the Vostok research station in Antarctica. As you have probably assumed, the depth of the ice core is related to how old the ice is; deep ice is older. There are two other variables that you will be analyzing from the ice cores. You will analyze CO2 concentration, which has been measured from air bubbles trapped in the ice. We can use these data to see what rates of change were like during this pre-historic period, during which human activity was minimal.

[Ice core data](https://data.ess-dive.lbl.gov/view/doi%3A10.3334%2FCDIAC%2FATG.009):

Vostok Core, back to 400,000 years before the present day

-   Description of data set: <https://data.ess-dive.lbl.gov/view/doi%3A10.3334%2FCDIAC%2FATG.009>
-   data: <https://data.ess-dive.lbl.gov/catalog/d1/mn/v2/object/ess-dive-457358fdc81d3a5-20180726T203952542>

You will use this code to download the data to your computer.

```{r}
download.file("https://data.ess-dive.lbl.gov/catalog/d1/mn/v2/object/ess-dive-457358fdc81d3a5-20180726T203952542",
              destfile = "vostok.icecore.co2")
```

You can click on the file in your Files pane to view it before reading into R.

**Question 10:**

The broad question is: how do recent atmospheric CO2 levels compare to historical levels?

Your answer to Question 10 is going to be a mix of code chunks and text that you put below in "Answer 10:"

-   Describe the data set: what are the columns and units? Where do the numbers come from?
-   What is the temporal resolution of the data?
-   Read in and prepare data for analysis.\
-   Reverse the ordering to create a chronological record so that each measurement is associcated with calender year.
-   Plot data.\
-   Combine this time series with the Mauna Loa data (the Warm-up exercise in this assignment). You will use "bind_rows()".
-   Plot the combined data. (the most recent time period must be on the right side of the plot).\
-   Describe your conclusions to the question "How do recent atmospheric CO2 levels compare to historical levels?" using your plot as supporting evidence.

**Answer 10:**

```{r}


```

# Render and committing

Remember to Render your document as HTML and comment+push to GitHub your code and rendered HTML that was created when you knitted the document. Your GitHub repository should have multiple commits with informative commit messages.

# Attribution

Include citation of any AI-generated assistance or discussion with classmates (per policy in syllabus). Proper documentation of AI-generated assistance includes the prompt, the source (e.g., ChatGPT), and the significant parts of the response. Proper documentation of discussion with classmates include listing their names and the components discussed.
